Finiteness, Cardinality, and Combinatorics in Homotopy Type Theory

### 2020-01-06 paper.tex
more

We will at this point take a moment to define some of the types we used to
define \(\mathcal{E}!\).
Lists, and membership thereof, are defined using \emph{containers}.
\begin{rm-definition}[Container] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a dependent
  pair of shapes, and positions indexed by those shapes. 
  \begin{equation}
    \mathbf{Container} = \sum_{(\text{shape} : \text{Type})} , \text{shape} \rightarrow \text{Type}
  \end{equation}
  We ``interpret'' a container into a pair of a specific shape, and a function
  which goes from positions to contents.
  \begin{equation}
    \llbracket S , P \rrbracket = \prod_{(X : \text{Type})} , \sum_{(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{rm-definition}
Containers can be used to define a wide variety of functors (streams, trees,
etc.): lists are all that interest us now.
\begin{rm-definition}[\mathbf{Fin}]
  \(\mathbf{Fin}\)(n) is the type of natural numbers smaller than \(n\).
  It is defined inductively, as follows:
  \begin{equation}
    \begin{aligned}
      \mathbf{Fin}(0) &= \bot \\
      \mathbf{Fin}(n + 1) &= \top + \mathbf{Fin}(n)
    \end{aligned}
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[Lists]
  The ``shape'' of lists is \(\mathbb{N}\), indicating the length of the list in
  question.
  \begin{equation}
    \mathbf{List} = \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket
  \end{equation}
\end{rm-definition}

@@

\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
\end{proof}
\subsection{Decidable Equality}
\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
\begin{rm-lemma}
  Any split enumerable type is a set.
\end{rm-lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma}
  \(\bot\) is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support list for \(\bot\) is the empty list.
  The cover proof is present via the principle of explosion.
\end{proof}
\begin{rm-lemma}
  \(\top\) is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support list for \(\top\) is the singleton list containing \(\text{tt}\).
  The cover proof is a constant function which returns an index into the first
  item in the list.
\end{proof}
Both of those cases were relatively simple.
Next, we will look into how to combine proofs of split enumerability.
\begin{rm-theorem}
  Split-enumerability is closed under \(\sum\).
\end{rm-theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \prod_{(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

@@

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsection{Equivalence}
Where split enumerability was the enumeration form of a surjection from
\(\mathbf{Fin}\), we see here that manifest Bishop finiteness is the enumeration
form of an \emph{equivalence} with \(\mathbf{Fin}\).
\begin{rm-lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{rm-lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  In other words, we can define a container-based version of bishop finiteness
  which is a reassociation of a contractible map, and then show that the
  container-based predicate is equivalent to the list-based.
\end{proof}
\subsection{Relationship to Split Enumerability}
We now show that manifest Bishop finiteness has equal strength to split
enumerability.
\begin{rm-lemma}
  Any manifest Bishop finite set is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support set carries over simply, and the cover proof can be taken from the
  first component of the cover proof from the proof of manifest Bishop
  finiteness.
\end{proof}
\begin{rm-theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{rm-theorem}
\begin{proof}
  Let \(E\) be a proof of split enumerability for some set \(A\).
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function \(\text{uniques}\) which filters
  out duplicates from lists of \(A\)s.
  \begin{equation}
    \text{uniques} : \textbf{List}(A) \rightarrow \textbf{List}(A)
  \end{equation}
  This gives us our support list.

@@

  The bulk of the proof now is concerned with proving the following:
  \begin{equation}
    \left( \prod_{(x : \mathbf{Fin}(n))} , \mathcal{E}!(A(x)) \right) \rightarrow \mathcal{E}!\left( \prod_{(x : \mathbf{Fin}(n))} , A(x) \right)
  \end{equation}
  Our strategy to accomplish this will be to consider functions from
  \(\mathbf{Fin}(n)\) as \(n\)-tuples over some type family \(T\).
  \begin{equation}
    \begin{aligned}
      \mathbf{Tuple}(T, 0)   &= \top \\
      \mathbf{Tuple}(T, n+1) &= A(0) \times \mathbf{Tuple}(T \circ \text{suc}, n)
    \end{aligned}
  \end{equation}
  This type is manifestly Bishop finite, as it is constructed only from products
  and the unit type.

@@

  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{pi-clos}
\end{proof}
\section{Manifest Enumerability} \label{manifest-enumerability}
As we said in section~\ref{split-enumerable}, we will now retrieve a distinction
between enumerability and manifest Bishop finiteness.
We do so by propositionally truncating the membership proof:
\begin{rm-definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) = \sum_{(\mathit{xs} : \mathbf{List}(A))} , \prod_{(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{rm-definition}
\subsection{Higher Homotopy Levels}
By hiding the position, we have essentially removed the ``decidable'' component
from split enumerability.
Our predicate now becomes general enough to work with non-sets: we will show
here that the circle is manifestly enumerable.
\begin{rm-theorem}
  The circle \(S^1\) is manifestly enumerable.
\end{rm-theorem}
\begin{proof}
  As the cover proof is a truncated proposition, we need only consider the point
  constructors, making this poof the same as the proof of split enumerability on
  \(\top\).

>>>>

We will at this point take a moment to define some of the types we used to
define \(\mathcal{E}!\).
Lists, and membership thereof, are defined using \emph{containers}.
\begin{rm-definition}[Container] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S \triangleright P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  \begin{equation}
    \begin{aligned}
      S : \text{Type}, P : S \rightarrow \text{Type} \\
      \mathbf{Container} = S \triangleright P
    \end{aligned}
  \end{equation}
  We ``interpret'' a container into a functor defined like so:
  \begin{equation}
    \llbracket S \triangleright P \rrbracket = \prod_{(X : \text{Type})} , \sum_{(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{rm-definition}
Containers can be used to define a wide variety of functors (streams, trees,
etc.): lists are all that interest us now.
\begin{rm-definition}[\mathbf{Fin}]
  \(\mathbf{Fin}\)(n) is the type of natural numbers smaller than \(n\).
  It is defined inductively, as follows:
  \begin{equation}
    \begin{aligned}
      \mathbf{Fin}(0) &= \bot \\
      \mathbf{Fin}(n + 1) &= \top + \mathbf{Fin}(n)
    \end{aligned}
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[Lists]
  The ``shape'' of lists is \(\mathbb{N}\), indicating the length of the list in
  question.
  \begin{equation}
    \mathbf{List} = \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket
  \end{equation}
\end{rm-definition}

@@

\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
\end{proof}
\subsection{Decidable Equality}
\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
\begin{rm-lemma}
  Any split enumerable type is a set.
\end{rm-lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma}
  \(\bot\), \(\top\), and \(\mathbf{Bool}\) are all split enumerable.
\end{rm-lemma}
\begin{proof}
  These non-recursive types have similar, simple proofs.
  For each we first provide the support list: they are \([]\), \([\text{tt}]\),
  and \([\text{false}, \text{true}]\) respectively.
  The cover proof should return an index which points at the given element: for
  \(\bot\) this function is present via the principle of explosion, for \(\top\)
  we always return a \(0\), and for \(\mathbf{Bool}\) we return a \(0\) for
  \(\text{false}\), and a \(1\) for \(\text{true}\).
  \begin{multicols}{3}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{unit}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{empty}
  \end{multicols}
\end{proof}
Next, we will look into how to combine proofs of split enumerability.
\begin{rm-theorem}
  Split-enumerability is closed under \(\sum\).
\end{rm-theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \prod_{(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

@@

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsection{Equivalence}
Where split enumerability was the enumeration form of a surjection from
\(\mathbf{Fin}\), we see here that manifest Bishop finiteness is the enumeration
form of an \emph{equivalence} with \(\mathbf{Fin}\).
\begin{rm-lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{rm-lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  In other words, the definition of Bishop finiteness is itself a reassociation
  of a contractible map.
\end{proof}
\subsection{Relationship to Split Enumerability}
We now show that manifest Bishop finiteness has equal strength to split
enumerability.
\begin{rm-lemma}
  Any manifest Bishop finite set is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support set carries over simply, and the cover proof can be taken from the
  first component of the cover proof from the proof of manifest Bishop
  finiteness.
\end{proof}
\begin{rm-theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{rm-theorem}
\begin{proof}
  Let \(E\) be a proof of split enumerability for some set \(A\).
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function \(\text{uniques}\) which filters
  out duplicates from lists of \(A\)s.
  \begin{equation}
    \text{uniques} : \textbf{List}(A) \rightarrow \textbf{List}(A)
  \end{equation}
  This gives us our support list.

@@

  The bulk of the proof now is concerned with proving the following:
  \begin{equation}
    \left( \prod_{(x : \mathbf{Fin}(n))} , \mathcal{E}!(A(x)) \right) \rightarrow \mathcal{E}!\left( \prod_{(x : \mathbf{Fin}(n))} , A(x) \right)
  \end{equation}
  Our strategy to accomplish this will be to consider functions from
  \(\mathbf{Fin}(n)\) as \(n\)-tuples over some type family \(T : \mathbb{N}
  \rightarrow \text{Type}\).
  \begin{equation}
    \begin{aligned}
      \mathbf{Tuple}(T, 0)   &= \top \\
      \mathbf{Tuple}(T, n+1) &= T(0) \times \mathbf{Tuple}(T \circ \text{suc}, n)
    \end{aligned}
  \end{equation}
  This type is manifestly Bishop finite, as it is constructed only from products
  and the unit type.

@@

  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{pi-clos}
\end{proof}
\section{Manifest Enumerability} \label{manifest-enumerability}
A defining feature of split enumerability and Bishop finiteness is decidability:
both predicates imply a decidable equality function on the underlying type.
To find a predicate for finiteness which doesn't imply this, and therefore
works with types other than just sets, we \emph{truncate} the membership proof.
\begin{rm-definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) = \sum_{(\mathit{xs} : \mathbf{List}(A))} , \prod_{(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{rm-definition}
\subsection{Propositional Truncation}
\begin{rm-definition}[Propositional Truncation]
  The type \(\lVert A \rVert\) on some type \(A\) is a propositionally truncated
  proof of \(A\).
  In other words, it is a proof that some \(A\) exists, but it does not tell you
  \emph{which} \(A\).

  It is defined as a Higher Inductive Type:
  \begin{equation}
    \begin{aligned}
      \lVert A \rVert &= \\
        &| \; \lvert \cdot \rvert : A \rightarrow \lVert A \rVert ; \\
        &| \; \text{squash} : \prod_{(x, y : \lVert A \rVert)} , x \equiv y  ; \\
    \end{aligned}
  \end{equation}

  To eliminate from \(\lVert A \rVert\) we need to prove that we don't ``care''
  about the specific choice of \(A\).
  In other words, given a function \(f : A \rightarrow B\), to construct a
  function \(\lVert A \rVert \rightarrow B\), \(f\) needs to satisfy the following:
  \begin{equation}
    \prod_{(x , y : A)} , f(x) \equiv f(y)
  \end{equation}
  Formally speaking, \(f\) needs to be ``coherently constant''
  \cite{krausGeneralUniversalProperty2015}.
\end{rm-definition}
\subsection{Higher Homotopy Levels}
By hiding the position, we have essentially removed the ``decidable'' component
from split enumerability.
Our predicate now becomes general enough to work with non-sets: we will show
here that the circle is manifestly enumerable.
\begin{rm-theorem}
  The circle \(S^1\) is manifestly enumerable.
\end{rm-theorem}
\begin{proof}
  As the cover proof is a truncated proposition, we need only consider the point
  constructors, making this poof the same as the proof of split enumerability on
  \(\top\).
